Defining Transcriptome using RNA Sequencing\\ of {\em Streptococcus mutans}

### 2012-01-23 package/manuscript.Rnw
Just an update.

\author{Sang Chul Choi$^{1}$, Charles D.\ Danko$^{1}$, 
Melissa J.\ Hubisz$^{1}$, \\
Paulina Dmitrievna Pavinski Bitar$^{2}$, \\
Robert A.\ Burne$^{3}$
Michael J.\ Stanhope$^{2}$, and Adam Siepel$^{1}$}

@@

@@

<<manuscript,echo=FALSE,results=hide>>=

@@

<<manuscript,eval=FALSE,echo=FALSE,results=hide>>=

@@

\fixme{We may need dot plots where x-axis represents fold changes, and y-axis
functional classes. Dots may be colored by p-values. This would show most of
information in one plot. We could label some of dots to emphasixe. Done?}
\fixme{Use KEGG for functional category study. Done?}
\fixme{Display results in UCSC genome browser. Done?}
\fixme{Use two factor analysis. Done?}

>>>>

\author{Sang Chul Choi$^{1}$, Charles D.\ Danko$^{1}$, 
Melissa J.\ Hubisz$^{1}$, \\
Paulina Dmitrievna Pavinski Bitar$^{2}$, \\
\fixme{Add more coauthors and change the order of authors if needed}, \\
Robert A.\ Burne$^{3}$
Michael J.\ Stanhope$^{2}$, and Adam Siepel$^{1}$}

@@

\fixme{Add corresponding authors if needed.}

@@

<<echo=FALSE,results=hide>>=

@@

<<eval=FALSE,echo=FALSE,results=hide>>=

@@

\fixme{We may need dot plots where x-axis represents fold changes, and y-axis
functional classes. Dots may be colored by p-values. This would show most of
information in one plot. We could label some of dots to emphasize. Done?}

% tableGlucoseOnly's functional classes can be multiple. We still need to have
% different colors using p-values.
% dotplot(tableGlucoseOnly[,"Functional Class"]~tableGlucoseOnly[,"fold"])

\fixme{Display results in UCSC genome browser. Done?}

\fixme{Young's GO analysis with the comparisons}
We used a q-value of $1\times 10^{-3}$ to determine differentially expressed
genes. Note that we had another q-value cutoff to list gene ontology terms for
functional category enrichment tests; 0.05 was the cutoff value for this.
We found gene ontology terms enriched with differentially expressed genes in the
comparison of UA159 and TW1 glucose-grown cells:
phosphotransferase system, 
glycogen biosynthetic process, 
sugar:hydrogen symporter activity, and
carbohydrate metabolic process (table \ref{tab:goseqGlucoseOnly}).
Because the number of differentially expressed genes in galactose-grown cells is
was smaller than that of glucose-grown cells, we found only one gene ontology
term, branched chain family amino acid biosynthetic process
(table \ref{tab:goseqGalactoseOnly}).
The comparisons between glucose-grown and galactose-grown UA159
(table \ref{tab:goseqUA159Only}), and that from TW1 mutant-type
(table \ref{tab:goseqTW1Only}) showed very similar functional categories. Ranks
of gene ontology terms were different. 
Lactose metabolic process was top in TW1, and
phosphoenolpyruvate-dependent sugar phosphotransferase system was top in UA159.

\fixme{Use KEGG for functional category study. Done?}
(table \ref{tab:kegseqGlucoseOnly})
(table \ref{tab:kegseqGalactoseOnly})
(table \ref{tab:kegseqUA159Only})
(table \ref{tab:kegseqTW1Only})

\fixme{Use two factor analysis. Done?}
Figure \ref{figClust}
Because the effects of nutrient changes appeared to be similar whether UA159
wild-type or TW1 mutant-type was used, we used all of the 
\Sexpr{length(colnames(counts(smutansGenes)))} samples to compare glucose-grown
and galactose grown cells.

<<echo=FALSE,results=hide>>=
allSample <- newSmutans( smutansGenes, 
                         title="All Samples for Comparing Conditions" )
allSample <- smutans.de2( allSample )
save(allSample, file=file.path("allSample.RData"))
#load("allSample.RData")
tabUA159onlyAndAllSample <- smutans.compareResult( ua159only, 
                                                   allSample )
#addmargins( tabUA159onlyAndAllSample )
@
Of the total \Sexpr{sum(tabUA159onlyAndAllSample)} gene,
\Sexpr{tabUA159onlyAndAllSample[2,2]} genes were found in both analyses. The
analysis using all of the sample found \Sexpr{tabUA159onlyAndAllSample[1,2]}
genes. The analysis using ua159-only sample found \Sexpr{tabUA159onlyAndAllSample[2,1]}
genes.  Figure \ref{figCompareUA159onlyAndAllSample} shows the raw p-values from
the two-factor analysis versus those from the single-factor analysis. The
p-values from the two-factor analysis are overall smaller than those from the
single-factor analysis. We would expect that this was the case because of higher
power of the two-factor analysis.

### 2012-01-23 package/supplementary.Rnw
Just an upate.
Just an update.

<<echo=FALSE,results=hide>>=
# not run when smutans R package is installed
# Use this when smutans R package is being tested.
library(DESeq)
library(GenomicRanges)
library(plotrix)
library(xtable)
library(RColorBrewer)
source("smutans/R/class_and_slots.R")
source("smutans/R/core.R")
source("smutans/R/methods.R")
load("smutans/data/smutansData.tx.RData")
load("smutans/data/smutansData.txGenes.RData")
load("smutans/data/smutansData.txPileup.RData")
load("smutans/data/smutans.feature.genes.RData")
load("smutans/data/smutans.genes.criteria.RData")
load("smutans/data/smutans.go.genes.RData")
load("smutans/data/smutans.cat.desc.RData")
load("smutans/data/smutansGenes.RData")
@

@@

<<echo=FALSE,results=tex>>=
#glucoseOnly <- newSmutans( smutansGenes, title="Glucose Only" )
#glucoseOnly <- smutans.de2( glucoseOnly, condition="glucose", 
#                            condA="ua159", condB="tw1" )
#save(glucoseOnly, file=file.path("glucoseOnly.RData"))
load("glucoseOnly.RData")
x <- smutans.de2List( glucoseOnly, qval=1e-3 )
x <- x[,c("id","baseMeanA","baseMeanB","log2FoldChange","padj")]
colnames(x) <- c("Gene ID","UA159","TW1","fold", "p-value")
x <- merge(x, geneDescription, by.x = "Gene ID", by.y = "Gene ID")
x.big <- xtable( x[order(x[,"p-value"]),],
                 display=c("s","s","d","d","f","e","s"),
                 digits=c(0,0,0,0,1,0,0),
                 align=c('l','l','r','r','r','r','p{3.5in}'),
                 label='tabGlucoseOnly',
                 caption='List of genes for the comparison UA159 and TW1 under glocuse-grown cells',
               )
print( x.big,
       tabular.environment='longtable',
       caption.placement="top",
       include.rownames=FALSE,
       floating=FALSE) 
x <- merge(x, functionalDescription, by.x = "Gene ID", by.y = "Gene ID")
@
\clearpage{}

We confirmed that the most upregulated genes in the TW1 mutant-type encoded the
components of the PDH enzyme complex (SMU.1421, SMU.1422, SMU.1423, and
SMU.1424).  
\Sexpr{length(grep("Energy metabolism",x[,"Functional Class"]))}
\Sexpr{length(grep("PTS",x[,"Functional Class"]))}
\Sexpr{length(grep("hypothetical",x[,"Gene Description"]))}

@@

@@

%=============================================================================
%\begin{figure}[!ht]
%\includegraphics[scale=0.42]{figures/clonalorigin}
%\caption[Illustration of recombinant tree]{
%{\bf Illustration of a recombinant tree in the ClonalOrigin model.}
%(A) A recombinant tree for our five-sample clonal frame with four
%recombinant edges (dashed lines with arrows running from donors to
%recipients).  (B) Genomic 
%regions associated with recombinant edges.  Recombination 
%intensities (number of edge types present) and gene tree labels are shown at
%bottom. (C)~Induced local gene trees.
%Bold lines indicate branches of the clonal frame, while dashed lines indicate
%recombinant edges. See section \textbf{Summary statistics of recombinant trees} in
%\textbf{Supplementary Text} for additional details.}
%\label{fig:clonalorigin}
%\end{figure}
%\clearpage{}%
%=============================================================================

>>>>

<<echo=FALSE,results=hide>>=
# not run when smutans R package is installed
# Use this when smutans R package is being tested.
library(DESeq)
library(goseq)
library(GenomicRanges)
library(plotrix)
library(xtable)
library(RColorBrewer)
source("smutans/R/class_and_slots.R")
source("smutans/R/core.R")
source("smutans/R/methods.R")
load("smutans/data/smutansData.tx.RData")
load("smutans/data/smutansData.txGenes.RData")
load("smutans/data/smutansData.txPileup.RData")
load("smutans/data/smutans.feature.genes.RData")
load("smutans/data/smutans.genes.criteria.RData")
load("smutans/data/smutans.go.genes.RData")
load("smutans/data/smutans.cat.desc.RData")
load("smutans/data/smutans.keg.genes.RData")
load("smutans/data/smutans.keg.desc.RData")
load("smutans/data/smutansGenes.RData")
@
<<eval=FALSE,echo=FALSE,results=hide>>=
# not run without smutans R package installed
# Remove eval=FALSE when the R package is installed
library(smutans)
load(smutansData.tx)
load(smutansData.txGenes)
load(smutansData.txPileup)
load(smutans.feature.genes)
data(smutans.genes.criteria)
data(smutans.go.genes)
data(smutans.cat.desc)
data(smutans.keg.genes)
data(smutans.keg.desc)
data(smutansGenes)
@

@@

<<echo=FALSE,results=tex>>=
#glucoseOnly <- newSmutans( smutansGenes, title="Glucose Only" )
#glucoseOnly <- smutans.de2( glucoseOnly, condition="glucose", 
#                            condA="ua159", condB="tw1" )
#save(glucoseOnly, file=file.path("glucoseOnly.RData"))
load("glucoseOnly.RData")
x <- smutans.de2List( glucoseOnly, qval=1e-3 )
x <- x[,c("id","baseMeanA","baseMeanB","log2FoldChange","padj")]
colnames(x) <- c("Gene ID","UA159","TW1","fold", "p-value")
x <- merge(x, geneDescription, by.x = "Gene ID", by.y = "Gene ID")
x.big <- xtable( x[order(x[,"p-value"]),],
                 display=c("s","s","d","d","f","e","s"),
                 digits=c(0,0,0,0,1,0,0),
                 align=c('l','l','r','r','r','r','p{3.5in}'),
                 label='tabGlucoseOnly',
                 caption='List of genes for the comparison UA159 and TW1 under glucose-grown cells',
               )
print( x.big,
       tabular.environment='longtable',
       caption.placement="top",
       include.rownames=FALSE,
       floating=FALSE) 
x <- merge(x, functionalDescription, by.x = "Gene ID", by.y = "Gene ID")
@
\clearpage{}

@@

\clearpage{}
%-----------------------------------------------------------------------------
\begin{table}[!h]
\caption{{\bf Gene Ontology (GO) enrichments for differentially expressed
genes in the comparison UA159 and TW1 grown in glucose.}}
\vspace{1ex}
\noindent \begin{centering}
\begin{tabular}{cl}
\hline 
$q^a$ & Description \\
\hline 
%-----------------------------------------------------------------------------
<<echo=FALSE,results=tex>>=
smutans.de2Goseq ( glucoseOnly, qval=1e-3, 
                   feature.gene=smutans.feature.genes,
                   go.genes=smutans.go.genes,
                   cat.desc=smutans.cat.desc )
@
%-----------------------------------------------------------------------------
\hline 
\end{tabular}
\par\end{centering}
\begin{flushleft}
$^a$False discovery rate estimated by the
Benjamini-Hochberg method.  All categories having at least ten genes and
$q\leq 0.05$ are displayed.\\ 
\end{flushleft}
\label{tab:goseqGlucoseOnly}
\end{table}
%-----------------------------------------------------------------------------

\clearpage{}
%-----------------------------------------------------------------------------
\begin{table}[!h]
\caption{{\bf Gene Ontology (GO) enrichments for differentially expressed
genes in the comparison UA159 and TW1 grown in galactose.}}
\vspace{1ex}
\noindent \begin{centering}
\begin{tabular}{cl}
\hline 
$q^a$ & Description \\
\hline 
%-----------------------------------------------------------------------------
<<echo=FALSE,results=tex>>=
smutans.de2Goseq ( galactoseOnly, qval=1e-3, 
                   feature.gene=smutans.feature.genes,
                   go.genes=smutans.go.genes,
                   cat.desc=smutans.cat.desc )
@
%-----------------------------------------------------------------------------
\hline 
\end{tabular}
\par\end{centering}
\begin{flushleft}
$^a$False discovery rate estimated by the
Benjamini-Hochberg method.  All categories having at least ten genes and
$q\leq 0.05$ are displayed.\\ 
\end{flushleft}
\label{tab:goseqGalactoseOnly}
\end{table}
%-----------------------------------------------------------------------------

\clearpage{}
%-----------------------------------------------------------------------------
\begin{table}[!h]
\caption{{\bf Gene Ontology (GO) enrichments for differentially expressed
genes in the comparison of glucose-grown and galactose-grown UA159 wild-type
cells.}}
\vspace{1ex}
\noindent \begin{centering}
\begin{tabular}{cl}
\hline 
$q^a$ & Description \\
\hline 
%-----------------------------------------------------------------------------
<<echo=FALSE,results=tex>>=
smutans.de2Goseq ( ua159only, qval=1e-3, 
                   feature.gene=smutans.feature.genes,
                   go.genes=smutans.go.genes,
                   cat.desc=smutans.cat.desc )
@
%-----------------------------------------------------------------------------
\hline 
\end{tabular}
\par\end{centering}
\begin{flushleft}
$^a$False discovery rate estimated by the
Benjamini-Hochberg method.  All categories having at least ten genes and
$q\leq 0.05$ are displayed.\\ 
\end{flushleft}
\label{tab:goseqUA159Only}
\end{table}
%-----------------------------------------------------------------------------

\clearpage{}
%-----------------------------------------------------------------------------
\begin{table}[!h]
\caption{{\bf Gene Ontology (GO) enrichments for differentially expressed
genes in the comparison of glucose-grown and galactose-grown TW1 mutant-type
cells.}}
\vspace{1ex}
\noindent \begin{centering}
\begin{tabular}{cl}
\hline 
$q^a$ & Description \\
\hline 
%-----------------------------------------------------------------------------
<<echo=FALSE,results=tex>>=
smutans.de2Goseq ( tw1only, qval=1e-3, 
                   feature.gene=smutans.feature.genes,
                   go.genes=smutans.go.genes,
                   cat.desc=smutans.cat.desc )
@
%-----------------------------------------------------------------------------
\hline 
\end{tabular}
\par\end{centering}
\begin{flushleft}
$^a$False discovery rate estimated by the
Benjamini-Hochberg method.  All categories having at least ten genes and
$q\leq 0.05$ are displayed.\\ 
\end{flushleft}
\label{tab:goseqTW1Only}
\end{table}
%-----------------------------------------------------------------------------

\clearpage{}
%-----------------------------------------------------------------------------
\begin{table}[!h]
\caption{{\bf KEGG enrichments for differentially expressed
genes in the comparison UA159 and TW1 grown in glucose.}}
\vspace{1ex}
\noindent \begin{centering}
\begin{tabular}{cl}
\hline 
$q^a$ & Description \\
\hline 
%-----------------------------------------------------------------------------
<<echo=FALSE,results=tex>>=
smutans.de2Goseq ( glucoseOnly, qval=1e-3, 
                   feature.gene=smutans.feature.genes,
                   go.genes=smutans.keg.genes[c(4,3)],
                   cat.desc=smutans.keg.desc)
@
%-----------------------------------------------------------------------------
\hline 
\end{tabular}
\par\end{centering}
\begin{flushleft}
$^a$False discovery rate estimated by the
Benjamini-Hochberg method.  All categories having at least ten genes and
$q\leq 0.05$ are displayed.\\ 
\end{flushleft}
\label{tab:kegseqGlucoseOnly}
\end{table}
%-----------------------------------------------------------------------------

\clearpage{}
%-----------------------------------------------------------------------------
\begin{table}[!h]
\caption{{\bf KEGG enrichments for differentially expressed
genes in the comparison UA159 and TW1 grown in galactose.}}
\vspace{1ex}
\noindent \begin{centering}
\begin{tabular}{cl}
\hline 
$q^a$ & Description \\
\hline 
%-----------------------------------------------------------------------------
<<echo=FALSE,results=tex>>=
smutans.de2Goseq ( galactoseOnly, qval=1e-3, 
                   feature.gene=smutans.feature.genes,
                   go.genes=smutans.keg.genes[c(4,3)],
                   cat.desc=smutans.keg.desc)
@
%-----------------------------------------------------------------------------
\hline 
\end{tabular}
\par\end{centering}
\begin{flushleft}
$^a$False discovery rate estimated by the
Benjamini-Hochberg method.  All categories having at least ten genes and
$q\leq 0.05$ are displayed.\\ 
\end{flushleft}
\label{tab:kegseqGalactoseOnly}
\end{table}
%-----------------------------------------------------------------------------

\clearpage{}
%-----------------------------------------------------------------------------
\begin{table}[!h]
\caption{{\bf KEGG enrichments for differentially expressed
genes in the comparison of glucose-grown and galactose-grown UA159 wild-type
cells.}}
\vspace{1ex}
\noindent \begin{centering}
\begin{tabular}{cl}
\hline 
$q^a$ & Description \\
\hline 
%-----------------------------------------------------------------------------
<<echo=FALSE,results=tex>>=
smutans.de2Goseq ( ua159only, qval=1e-3, 
                   feature.gene=smutans.feature.genes,
                   go.genes=smutans.keg.genes[c(4,3)],
                   cat.desc=smutans.keg.desc)
@
%-----------------------------------------------------------------------------
\hline 
\end{tabular}
\par\end{centering}
\begin{flushleft}
$^a$False discovery rate estimated by the
Benjamini-Hochberg method.  All categories having at least ten genes and
$q\leq 0.05$ are displayed.\\ 
\end{flushleft}
\label{tab:kegseqUA159Only}
\end{table}
%-----------------------------------------------------------------------------

\clearpage{}
%-----------------------------------------------------------------------------
\begin{table}[!h]
\caption{{\bf KEGG enrichments for differentially expressed
genes in the comparison of glucose-grown and galactose-grown TW1 mutant-type
cells.}}
\vspace{1ex}
\noindent \begin{centering}
\begin{tabular}{cl}
\hline 
$q^a$ & Description \\
\hline 
%-----------------------------------------------------------------------------
<<echo=FALSE,results=tex>>=
smutans.de2Goseq ( tw1only, qval=1e-3, 
                   feature.gene=smutans.feature.genes,
                   go.genes=smutans.keg.genes[c(4,3)],
                   cat.desc=smutans.keg.desc)
@
%-----------------------------------------------------------------------------
\hline 
\end{tabular}
\par\end{centering}
\begin{flushleft}
$^a$False discovery rate estimated by the
Benjamini-Hochberg method.  All categories having at least ten genes and
$q\leq 0.05$ are displayed.\\ 
\end{flushleft}
\label{tab:kegseqTW1Only}
\end{table}
%-----------------------------------------------------------------------------

% END OF TABLE

@@

\begin{figure}
\centering
\includegraphics[width=.9\textwidth]{supplementary-figClust}
\caption{Heatmap showing the Euclidean distances between the samples
  as calculated from the variance-stabilizing transformation of the
  count data.}
\label{figClust}
\end{figure}
<<figClust,echo=FALSE,result=hide,fig=TRUE>>=
allSample <- newSmutans( smutansGenes )
smutans.de2Clust( allSample )
@
